Kleene Categories

### 2025-07-15 kleenecat/kleenecat.tex
progress on distributive law

@@

\begin{figure}
  \begin{align*}
    x + (y + z) &= (x + y) + z & x + y &= y + x\\
    x + 0 &= x & x + x &= x\\
    x(yz) &= (xy)z & x1 &= 1x = x\\
    x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
    x0 &= 0x = x & & \\
    1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
     b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
  \end{align*}
  \label{fig:axioms}
  \caption{Kleene algebra axioms}
\end{figure}

@@

It has been observed that Brzozowski's derivative can be more conceptually
explained as a distributive law between the Kleene algebra monad and the
DFA functor $2 \times (\Sigma \Rightarrow -)$.

@@

Since the axiomatization of Kleene algebras over an alphabet $\Sigma$
is a Horn theory, the category of Kleene algebras and Kleene algebra
morphisms is monadic over $\Set$. Let $E_\Sigma : \Set \to \Set$. be the 
Kleene algebra monad. It is possible to show the following theorem.

\begin{theorem}
  There is a distributive law $E_\Sigma(2 \times (\Sigma \Rightarrow X)) \to 2 \times (\Sigma \Rightarrow E_\Sigma(X))$.
\end{theorem}
\begin{proof}
  The component $E_\Sigma(2 \times (\Sigma \Rightarrow X)) \to 2$ is
  defined using the universal property of free algebraic structures,
  i.e. it suffices to define a morphism $2 \times (\Sigma \Rightarrow
  X)) \to 2$, which we choose it to be the first projection
  $\pi_1$. The component $E_\Sigma(2 \times (\Sigma \Rightarrow X))
  \to (\Sigma \Rightarrow E_\Sigma(X))$ requires us to be more
  clever. There is a ``non-standard'' Kleene algebra structure on
  $(\Sigma \Rightarrow E_\Sigma(X)) \times E_\Sigma(X)$ that makes
  this possible.
\end{proof}

@@

We now want to prove that the distributive law above generalizes to
the Kleene category case. We begin by showing that the ``non-standard''
structure defined in the previous section can be extended to the
proof-relevant setting.

\begin{lemma}
  Let $\cat{K}$ and $\cat{L}$ be two Kleene categories, $F : \cat{L}
  \to \cat{K}$ and $G : \cat{L} \to \cat{K}$ two Kleene category
  morphisms. We can equip $\cat{K} \times \cat{L}$ with the following
  Kleene category structure:
  \begin{itemize}
  \item $(A, X) + (B, Y) = (A + B, X + Y)$
  \item $0 = (0, 0)$
  \item $(A, X) \otimes (B, Y) = (A \otimes F(Y) + G(X) \otimes B, X \otimes Y)$
  \item $(A, X)^* = (A \otimes F(X^*), X^*)$
  \end{itemize}
\end{lemma}
%% \begin{proof}
%%   Since the morphism structure is defined componentwise, it is
%%   immediate to show that the proposed cocartesian structure is
%%   valid. For the monoidal structure, the bifunctoriality of $\otimes$
%%   follows from the bifunctoriality of the monoidal structures
%%   $\otimes$ and $+$ in $\cat{K}$. Next, we define the coherence
%%   isomorphisms. As suggested by the action on the $\cat{L}$ component,
%%   such isomorphisms are the same as in $\cat{L}$. The $\cat{K}$
%%   components of these morphisms are more interesting:
%% \end{proof}

With this lemma and the 2-monadicity of Kleene categories over $\cat{Cat}$,
we can define the distributive law between the proof-relevant
DFA functor and the Kleene category 2-monad.

@@

Then we can prove the following
\begin{theorem}{Gluing for Kleene categories}
  Let $\cat K, \cat L, \cat M$ be Kleene categories and $\Phi : \cat K
  \to \cat M$ be a Kleene functor and $G : \cat L \to \cat M$ be a lax
  monoidal functor. Then $\Phi \downarrow G$ is a Kleene category, and
  the projections are (strict) Kleene functors.
\end{theorem}
\begin{proof}
  The monoidal and coproduct aspects of Kleene categories are handled
  by the previous lemma. We need to show that the monoidal structure
  is distributive and construct the Kleene star.

>>>>

\usepackage{mdframed}
\usepackage{xcolor}
\usepackage{tcolorbox}
\newcommand\makenote[4]{\par\noindent\begin{tcolorbox}
[size=small,colback=#2]\color{#3}\textbf{#4: }#1\end{tcolorbox}\noindent}
\definecolor{cornellred}{RGB}{196,18,48}
\definecolor{umichblue}{HTML}{00274C}

\newcommand\pa[1]{\makenote{#1}{cornellred}{white}{PA}}
\newcommand\msn[1]{\makenote{#1}{umichblue}{white}{MSN}}

@@

\begin{figure}
  \begin{align*}
    x + (y + z) &= (x + y) + z & x + y &= y + x\\
    x + 0 &= x & x + x &= x\\
    x(yz) &= (xy)z & x1 &= 1x = x\\
    x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
    x0 &= 0x = 0 & & \\
    1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
     b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
  \end{align*}
  \label{fig:axioms}
  \caption{Kleene algebra axioms}
\end{figure}

@@

\begin{definition}
  A distributive law between a monad $T$ and an endofunctor $T$ is a
  natural transformation $\lambda : TF \to F T$ making the following
  diagrams commute:
% https://q.uiver.app/#q=WzAsOCxbMSwwLCJGIl0sWzAsMSwiVEYiXSxbMiwxLCJGVCJdLFswLDIsIlRURiJdLFsyLDIsIkZUVCJdLFsxLDIsIlRGVCJdLFswLDMsIlRGIl0sWzIsMywiRlQiXSxbMSwyLCJcXGxhbWJkYSIsMl0sWzAsMSwiXFxldGFfRiIsMl0sWzAsMiwiRlxcZXRhIl0sWzMsNiwiXFxtdV9GIiwyXSxbNiw3LCJcXGxhbWJkYSJdLFszLDUsIlRcXGxhbWJkYSJdLFs1LDQsIlxcbGFtYmRhIFQiXSxbNCw3LCJGIFxcbXUiXV0=
\[\begin{tikzcd}
	& F \\
	TF && FT \\
	TTF & TFT & FTT \\
	TF && FT
	\arrow["{\eta_F}"', from=1-2, to=2-1]
	\arrow["{F\eta}", from=1-2, to=2-3]
	\arrow["\lambda"', from=2-1, to=2-3]
	\arrow["{T\lambda}", from=3-1, to=3-2]
	\arrow["{\mu_F}"', from=3-1, to=4-1]
	\arrow["{\lambda T}", from=3-2, to=3-3]
	\arrow["{F \mu}", from=3-3, to=4-3]
	\arrow["\lambda", from=4-1, to=4-3]
\end{tikzcd}\]
\end{definition}

It has been observed that Brzozowski's derivative can be more conceptually
explained as a distributive law between the Kleene algebra monad and the
DFA functor $2 \times (-)^\Sigma$.

@@

\pa{I can't help but think that there's more to the construction
below... It feels like $K^\Sigma \times K$ is ``fibered'' over $K$}

\begin{definition}
  \label{def:nonstandard}
  If $K$ is a Kleene algebra over $\Sigma$, we can equip
  $K^\Sigma \times K$ with the following Kleene algebra
  structure.
    \begin{align*}
    (f_1, e_1) + (f_2, e_2) &= (f_1 + f_2, e_1 + e_2)\\
    0 &= (\lambda c.\, 0, 0)\\
    (f_1, e_1) \cdot (f_2, e_2) &= (\lambda c.\, f_1(c)\cdot e_2 + \alpha(e_1)\cdot f_2(c), e_1\cdot e_2)\\
    1 &= (\lambda c.\, 0, 1)\\
    (f, e)^* &= (\lambda c.\, f(c)\cdot e^*, e^*)
    \end{align*}
\end{definition}

The proof that this is a Kleene algebra is boring, but unsurprising.
Because the axiomatization of Kleene algebras over an alphabet $\Sigma$
is a Horn theory, the category of Kleene algebras and Kleene algebra
morphisms is monadic over $\Set$. Let $E_\Sigma : \Set \to \Set$ be the 
Kleene algebra monad.

It is possible to show that there is a distributive law
$E_\Sigma(2 \times X^\Sigma) \to 2 \times (E_\Sigma(X)^\Sigma)$.
Its recursive definition is depicted in Figure~\ref{fig:distlaw}. 
%
\begin{figure}
  \begin{align*}
    \alpha : E_\Sigma(2 \times X^\Sigma) \to 2 \quad\quad & \delta : E_\Sigma(2 \times X^\Sigma) \to E_\Sigma(X)^\Sigma\\
  \alpha(e_1 + e_2) = \alpha(e_1) + \alpha(e_2) \quad \quad & \delta_c(e_1 + e_2) = \delta_c(e_1) + \delta_c(e_2)\\ 
  \alpha(e_1\cdot e_2) = \alpha(e_1) \land \alpha(e_2) \quad \quad& \delta_c(e_1\cdot e_2) = \delta_c(e_1)\cdot e_2 + \alpha(e_1)\cdot \delta_c(e_2)\\ 
  \alpha(e^*) = 1 \quad \quad& \delta_c(e^*) = \delta_c(e)e^*\\ 
  \alpha(1) = 1 \quad \quad& \delta_c(1) = 0\\ 
  \alpha(0) = 0 \quad \quad& \delta_c(0) = 0\\ 
  \alpha(d : \Sigma) = 0 \quad \quad& \delta_c(d : \Sigma) = [c = d]\\ 
  \alpha(b, f) = b \quad \quad& \delta_c(b, f) = f(c)  
\end{align*}
\caption{Distributive law for Kleene algebras}
\label{fig:distlaw}
\end{figure}

\begin{theorem}
  There is a distributive law $E_\Sigma(2 \times (X^\Sigma)) \to 2 \times (E_\Sigma(X)^\Sigma)$.
\end{theorem}
\begin{proof}
  The component $E_\Sigma(2 \times (X^\Sigma)) \to 2$ is
  defined using the universal property of free algebraic structures,
  i.e. it suffices to define a morphism $2 \times X^\Sigma \to 2$,
  which we choose it to be the first projection
  $\pi_1$. The component $E_\Sigma(2 \times (X^\Sigma)) \to (E_\Sigma(X)^\Sigma)$
  requires us to be more
  clever. Instantiating the ``non-standard'' Kleene algebra structure from
  Definition~\ref{def:nonstandard} to $E_\Sigma(X)$
  gives us a morphism $\psi : E_\Sigma(2 \times X^\Sigma) \to E_\Sigma(X)^\Sigma \times E_\Sigma(X)$
  by using the universal property of free Kleene
  algebras and lifting the function
  \[\lambda (b, f).\, (\lambda c.\, \eta(f(c)), 1) : 2 \times X^\Sigma \to E_\Sigma(X)^\Sigma \times E_\Sigma(X)\]
  The derivative $\delta$ is the postcomposition of $\psi$ with the
  projection function.
\end{proof}

@@

We now want to prove that the distributive law above generalizes to
the Kleene category case. First, we have to define a natural
transformation $K(\Set \times (-)^\Sigma) \to \Set \times (K -
)^\Sigma$ making the appropriate diagrams commute. We decompose the
natural transformation into two components.  The component $K(\Set
\times (-)^\Sigma) \to \Set$ is by equipping $\Set$ with a Kleene
category structure and then using the universal property of the free
Kleene category. The Kleene category structure on $\Set$ is the usual
(co)cartesian structure and $X^*$ are the lists on $X$. Therefore, it
suffices to define a natural transformation $\Set \times (-)^\Sigma \to \Set$,
which we define as the first projection.

For the component $K(\Set \times (-)^\Sigma) \to (K - )^\Sigma$,
we will once again use the universal property of free Kleene
categories. We begin by showing that the ``non-standard''
structure defined in the previous section can be extended to the
proof-relevant setting.

\pa{This is where a better understanding of what this ``non-standard''
  construction is actually doing could come in handy. Showing that the
  monoidal product satisfies all of the coherence diagrams is really gnarly.}
\begin{lemma}
  Let $\cat{K}$ and $\cat{L}$ be two Kleene categories, $F : \cat{L} \to \cat{K}$
  and $G : \cat{L} \to \cat{K}$ two Kleene functors.
  We can equip $\cat{K} \times \cat{L}$ with the following Kleene
  category structure:
  \begin{itemize}
  \item $(A, X) + (B, Y) = (A + B, X + Y)$
  \item $0 = (0, 0)$
  \item $(A, X) \otimes (B, Y) = (A \otimes F(Y) + G(X) \otimes B, X \otimes Y)$
  \item $I = (0, I)$
  \item $(A, X)^* = (A \otimes F(X^*), X^*)$
  \end{itemize}
\end{lemma}

Therefore we equip $K(\cat{C})^\Sigma \times K(\cat{C})$ with the
Kleene category structure given by the lemma above.
%% \begin{proof}
%%   Since the morphism structure is defined componentwise, it is
%%   immediate to show that the proposed cocartesian structure is
%%   valid. For the monoidal structure, the bifunctoriality of $\otimes$
%%   follows from the bifunctoriality of the monoidal structures
%%   $\otimes$ and $+$ in $\cat{K}$. Next, we define the coherence
%%   isomorphisms. As suggested by the action on the $\cat{L}$ component,
%%   such isomorphisms are the same as in $\cat{L}$. The $\cat{K}$
%%   components of these morphisms are more interesting:
%% \end{proof}

@@

Then we can prove the following
\begin{theorem}[Gluing for Kleene categories]
  Let $\cat K, \cat L, \cat M$ be Kleene categories and $\Phi : \cat K
  \to \cat M$ be a Kleene functor and $G : \cat L \to \cat M$ be a lax
  monoidal functor. Then $\Phi \downarrow G$ is a Kleene category, and
  the projections are (strict) Kleene functors.
\end{theorem}
\begin{proof}
  The monoidal and coproduct aspects of Kleene categories are handled
  by the previous lemma. We need to show that the monoidal structure
  is distributive and construct the Kleene star.
